---
title: 'Homework #9 Math 160'
author: "Ethan Ashby"
output: 
  html_document:
    css: ./homework-style.css
---

\newcommand{\ind}{\mathbb{I}}
\newcommand{\unifdist}{\textsf{Unif}}
\newcommand{\binomialdist}{\textsf{Bin}}
\newcommand{\geodist}{\textsf{Geo}}
\newcommand{\berndist}{\textsf{Bern}}
\newcommand{\gammadist}{\textsf{Gamma}}
\newcommand{\expdist}{\textsf{Exp}}
\newcommand{\poisdist}{\textsf{Pois}}
\newcommand{\tdist}{\textsf{t}}
\newcommand{\betadist}{\textsf{Beta}}
\newcommand{\negativebinomialdist}{\textsf{NegBin}}
\newcommand{\normaldist}{\textsf{N}}
\newcommand{\mean}{\mathbb{E}}
\newcommand{\prob}{\mathbb{P}}
\newcommand{\real}{\mathbb{R}}
\newcommand{\argmax}{\text{arg}\,\text{max}}
\newcommand{\cdf}{\operatorname{cdf}}
\newcommand{\median}{\operatorname{median}}
\newcommand{\conf}{\operatorname{conf}}
\newcommand{\floor}[1]{\lfloor{#1}\rfloor}
\newcommand{\ceil}[1]{\lceil{#1}\rceil}
\newcommand{\asin}{\operatorname{asin}}
\newcommand{\acos}{\operatorname{acos}}
\newcommand{\mrrtt}{\textsf{M}\textsf{R}^2\textsf{T}^2}

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(knitr)
```



```{r setup, include=FALSE}
opts_chunk$set(echo = TRUE, fig.height = 3, fig.width = 5)
#Color Format
colFmt <- function(x,color){
  outputFormat = opts_knit$get("rmarkdown.pandoc.to")
  if(outputFormat == 'latex')
    paste("\\textcolor{",color,"}{",x,"}",sep="")
  else if(outputFormat == 'html')
    paste("<font color='",color,"'>",x,"</font>",sep="")
  else
    x
}
#Box Format
boxFmt <- function(x){
  outputFormat = opts_knit$get("rmarkdown.pandoc.to")
  if(outputFormat == 'latex')
    paste("\\framebox{\\hspace*{1em} ", x,"}")
  else if(outputFormat == 'html')
    paste("<span class=\"boxed\">",x,"</span>")
  else
    x
}
boxed <- function(x) boxFmt(x)
```

```{r, echo = FALSE}
due_dateSpring2021 <- function(number) switch(number, "", "", "", "", "", "", "31 Mar, 2021", "7 Apr, 2021", "14 Apr, 2021", "21 Apr, 2021", "5 May, 2021")
```


Due:  `r due_dateSpring2021(9)`

1) Create solutions using `:::: {.solution}` and `::::` in the .Rmd file.

2) If the .css file is not working for you, type out the Problem headings using `# Problem 1`, `# Solution`, `# Problem 2`, and so on.

2) Knit to an .html file.

3) Print out the .html file inside a browser (such as Chrome or Safari) to a .pdf file.

4) Upload the .pdf file to Gradescope.

5) Mark the pages that your questions are on.

6) Be sure to box your answers with the `\boxed{put in box}` $\LaTeX$ command.

7) Be sure to give nonintegral answers to four significant digits.

```{r, echo = FALSE}
set.seed(1234567)
```

::::: {.problem}
Suppose \( \prob(A_1) = \prob(A_2) = \prob(A_3) = 0.15 \).  Give an upper bound on \( \prob(A_1 \cup A_2 \cup A_3) \).
:::::

::::{.solution}
If \( \prob(A_1) = \prob(A_2) = \prob(A_3) = 0.15 \), we can find the upper bound on \( \prob(A_1 \cup A_2 \cup A_3) \) by supposing that $A_1, A_2, A_3$ are all disjoint. So the upper bound on \( \prob(A_1 \cup A_2 \cup A_3)=3 \cdot 0.15 = \boxed{0.45}\).
::::

::::: {.problem}
For \( x \in \{0, 1\}^n \), consider the partial order where
\[
(x \preceq w) \leftrightarrow (\forall i \in \{1, \ldots, n\})(x(i) \leq w(i)).
\]

a.  What state \( u \) satisfies \( x \preceq u \) for all \( x \in \{0, 1\}^n \)?

b.  What state \( \ell \) satisfies \( \ell \preceq x \) for all \( x \in \{0, 1\}^n \)?
:::::

::::{.solution}
a.
If $x \preceq u$, this implies that $(\forall i \in \{1, \ldots, n\})(x(i) \leq u(i))$. Since $x \in \{0, 1\}^n$, $x(i) \in \{0, 1\}$, $\sup(x(i))=1$. In order for $x(i) \leq u(i)$ for all $i$, $\boxed{u=\{1\}^n}$, i.e., a vector of `s of length n.

b.
If $\ell \preceq x$, this implies that $(\forall i \in \{1, \ldots, n\})(\ell(i) \leq x(i))$. Since $x \in \{0, 1\}^n$, $x(i) \in \{0, 1\}$, $\inf(x(i))=0$. In order for $\ell(i) \leq x(i)$ for all $i$, $\boxed{\ell=\{0\}^n}$, i.e., a vector of 0s of length n.
::::

::::: {.problem}
This problem uses the `tidygraph`, `igraph`, and `ggraph` libraries.
```{r, warning = FALSE, message = FALSE}
library(tidygraph)
library(ggraph)
library(igraph)
```

This allows us to set up graphs such as the 4 by 4 lattice.
```{r, fig.width = 3, fig.height = 3}
g <- create_lattice(c(4, 4))
plot_lattice <- function(g) {
  p <- ggraph(g, layout = layout_on_grid(g)) +
    geom_edge_link() +
    geom_node_point(size = 12, shape = 21, color = "black", fill = "steelblue") +
    coord_fixed(clip = "off") +
    theme_graph()
  return(p)
}
plot_lattice(g)
```

Consider the following Gibbs step for the Ising model, which is monotonic when \( \beta > 0 \).

```{r}
step_local_gibbs <- function(x_n, i, u, beta) {
  n_0 = sum(x_n == 0)
  n_1 = sum(x_n == 1)
  return(as.integer(u < (exp(beta * n_1) / (exp(beta * n_1) + exp(beta * n_0)))))
}
```

Then the following code uses CFTP to generate draws exactly from the Ising model.

```{r}
cftp_local_gibbs <- function(n = 1, g, beta = 1) {
  steps <- 2^n
  num_nodes <- g %>% activate(nodes) %>% as_tibble() %>% nrow()
  i <- floor(num_nodes * runif(steps)) + 1
  u <- runif(steps)
  x <- rep(0, num_nodes)
  w <- rep(1, num_nodes)
  for (t in 1:steps) {
    neigh <- neighbors(g, i[t])
    x_n <- x[neigh]
    x[i[t]] <- step_local_gibbs(x_n, i[t], u[t], beta)
    w_n <- w[neigh]
    w[i[t]] <- step_local_gibbs(w_n, i[t], u[t], beta)
  }
  if (sum(x) == sum(w) | n > 10) {
    return(x)
  }
  else {
    y <- cftp_local_gibbs(n + 1, g, beta)
    for (t in 1:steps) {
      y_n <- y[neighbors(g, i[t])]
      y[i[t]] <- step_local_gibbs(y_n, i[t], u[t], beta)
    }
    return(y)
  }
}
```

Then the following calculates \( h(x) \) for a configuration on a graph.
```{r}
h <- function(x, g) {
  from <- g %>% activate(edges) %>% pull(from)
  to <- g %>% activate(edges) %>% pull(to)
  return(sum(x[from] == x[to]))
}
```

Using this algorithm to create \( 10 \) iid draws from the Ising model over a 4 by 4 square lattice at \( \beta = 1\) and estimate \( \mean(h(X)). \)  Report your estimate as \( a \pm b \).
:::::

::::{.solution}

Generate 10 values, estimate the result
```{r}
set.seed(1234567)
res<-replicate(10, cftp_local_gibbs(n=1, g, beta=1))
stats<-apply(res, MARGIN=2, FUN=h, g=g)

tibble(
  mean=mean(stats),
  se=sd(stats)/sqrt(length(stats))
) %>% kable()
```

My estimate for $\mean(h(X))=\boxed{18.60 \pm 0.78}$.
::::

::::: {.problem}
Continuing the last problem, use your algorithm to run a Markov chain 10 times starting at a stationary state \( Y_0 \) and using \( Y_0, \ldots, Y_{100} \) to estimate \( \mean(h(X)) \) on a 4 by 4 square lattice.  Report your estimate as \( a \pm b \).
:::::

::::{.solution}

```{r, echo = FALSE}
set.seed(1234567)

run_mchain_p5<-function(steps, gr=g, b=1){
  #data gathering vector
  res<-rep(0, steps)
  #obtain stationary state
  y<-cftp_local_gibbs(n=1, g=gr, beta=b)
  i<-floor(16*runif(steps))+1
  u<-runif(steps)
  for(j in 1:steps){
    y_n<-y[neighbors(gr, i[j])]
    y[i[j]] <- step_local_gibbs(y_n, i[j], u[j], beta=b)
    #produce new state y
    res[j]<-h(y, g=gr)
  }
  return(res)
}

res<-replicate(10, run_mchain_p5(steps=100))

tibble(
  mean=mean(colMeans(res)),
  sd=sd(colMeans(res))/sqrt(length(colMeans(res)))
) %>% kable()
```
My estimate of $\mean(h(X))=\boxed{19.56 \pm 0.61}$.

::::

::::: {.problem}
Write code to draw five points uniformly from \( [0, 10]^2 \) such that at least one is within distance 1.2 of the point \( (2.1, 3.5) \) using Bonferroni AR.
:::::

::::{.solution}
```{r}
set.seed(1234567)
#ar alg
bonferroni_ar_p5 <- function() { 
  a <- FALSE
  while (!a) {
    #choose a point
    i <- floor(5 * runif(1)) + 1
    
    #determine location of ith point
    r <- sqrt(runif(1.2))
    theta <- 2 * pi * runif(1)
    points <- matrix(rep(0, 10), ncol = 2) # set up points matrix 
    
    for (j in 1:5) {
      if (j == i)
        points[j,] <- c(2.1 + r * cos(theta), 3.5 + r * sin(theta))
      else
        points[j,] <- 10 * runif(2) }
    
    u <- runif(1)
    # Count how many points are within distance 1 of (3, 5)
    m <- sum(sapply(1:5, function(j) (sqrt(sum((points[j,] - c(2.1, 3.5))^2)) <= 1.2)))
    a <- (u < 1 / m) 
  }
return(points) 
}

res<-bonferroni_ar_p5() %>% as.data.frame()
ggplot()+geom_point(data=res, aes(x=V1, y=V2))+xlim(c(0,10))+ylim(c(0,10))+coord_fixed()+geom_polygon(data=NULL, aes(x=2.1+1.2*cos(seq(0,2*pi,length=100)), y=3.5+1.2*sin(seq(0,2*pi,length=100))), alpha=0.3, fill="red")+theme_minimal()
```
Bingo! Code is working!
::::


::::: {.problem}
Suppose that each of \( B_1, B_2, B_3, B_4 \) is equally likely to be 0 or 1.  Suppose further that the goal is to sample from the \( B_i \) conditioned on event \( S \) being true, where 
\[
S = (B_1 = 1 \vee B_2 = 1 \vee B_3 = 1) \wedge (B_2 = 0 \vee B_3 = 1 \vee B_4 = 0) \wedge (B_1 = 0 \vee B_2 = 0 \vee B_4 = 1).
\]

Write basic AR code to sample from this distribution.  Use 1000 samples to estimate \( \prob(B_2 = 1 \mid S). \)  Report your result as \( a \pm b \).
:::::

::::{.solution}
```{r}
set.seed(1234567)
basic_ar_p6<-function(){
  s<-FALSE
  while(!s){
    #draw b values
    b<-round(runif(4))
    s=(b[1]==1 | b[2]==1 | b[3]==1) & (b[2]==0 | b[3]==1 | b[4]==1) & (b[1]==0 | b[2]==0 | b[4]==1)
  }
  return(b[2])
}

res<-replicate(1000, basic_ar_p6())

tibble(
  mean=mean(res),
  se=sd(res)/sqrt(length(res))
) %>% kable()
```

My basic AR estimate of \( \prob(B_2 = 1 \mid S)= \boxed{0.478 \pm 0.016}\).
::::

::::: {.problem}
Consider state space \( \Omega = \{\{0, 1\}^4 : S\}. \)  Write code that takes an \( \mrrtt \) step where the input is the current state \( x \) and \( i \in \{1, 2, 3, 4\} \) with \( y[-i] = x[-i] \) and \( y[i] = 1 - x[i] \).
:::::

::::{.solution}
Note that if proposed move is in $S$, we will **always** accept the proposed move, because the distribution is uniform over $\Omega$. In other words, given they are both in $S$, the proposed state and current state have equal likelihoods, so $U \leq 1$ is always satisfied.

```{r}
mr2t2_step_p7<-function(x){
  i=floor(4*runif(1))+1
  # do step
  y<-x
  y[i]<-1-x[i]
  #note that the step will always happen
  s=(y[1]==1 | y[2]==1 | y[3]==1) & (y[2]==0 | y[3]==1 | y[4]==1) & (y[1]==0 | y[2]==0 | y[4]==1)
  if(s){
    return(y)
  }
  else{
    return(x)
  }
}
```

This step function returns the current state if the proposed state is not in $S$, and returns the proposed state if it is in $S$.
::::

::::: {.problem}
Consider two circles that overlap.  Circle \( A \) has area \( (1/2)\tau \cdot 1.1^2\).  Circle \( B \) has area \( (1 / 2)\tau 1.4^2 \).  There is a rectangle \( R \) of area 16 that contains both circles.  Let \( X \sim \unifdist(R) \).

a. What is \( \prob(X \in A) \)?

b. What is \( \prob(X \in B) \)?
:::::

::::{.solution}

a.
\( \prob(X \in A) = \frac{m(A)}{m(R)}=\frac{(1/2)\tau \cdot 1.1^2}{16}= \boxed{0.238}\)

b.
\( \prob(X \in B) = \frac{m(B)}{m(R)}=\frac{(1/2)\tau \cdot 1.4^2}{16}= \boxed{0.385}\)
::::

::::: {.problem}
Suppose the following code can be used to determine if a point is in \( A \) or \( B \), and to draw from the areas of circles \( A \) and \( B \).

```{r}
is_in_A <- function(x) return(sum((x - c(1.5, 1.2))^2) <= 1.1^2)
is_in_B <- function(x) return(sum((x - c(2.6, 1.8))^2) <= 1.4^2)
rA <- function() {
  theta <- 2 * pi * runif(1)
  r <- 1.1 * sqrt(runif(1))
  return(c(1.5 + r * cos(theta), 1.2 + r * sin(theta)))
}
rB <- function() {
  theta <- 2 * pi * runif(1)
  r <- 1.4 * sqrt(runif(1))
  return(c(2.6 + r*cos(theta), 1.8 + r*sin(theta)))
}
```

a. Write Bonferroni AR code using these functions to draw uniformly from \( A \cup B \).

b. Use your code to estimate the area of \( A \cap B \) divided by the area of \( A \cup B \) using 1000 samples.  Report your result as \( a \pm b \).
:::::

::::{.solution}
a.
```{r}
set.seed(1234567)

bonf_ar_p9<-function(){
  a<-FALSE
  while(!a){
    #draw an index according to the probabilities of 
    i<-sample(x=c(1,2,3), size=1, prob=c(pi*1.1^2 * 1/16, pi*1.4^2 * 1/16, 1-(pi*1.1^2 * 1/16 + pi*1.4^2 * 1/16)))
    u<-runif(1)
    if(i==1){
      x<-rA()
      #we don't want to count the intersection twice, so we only
      #pass the TRUE flag our uniform is less than the number of conditions passed
      a<-(u <= 1/(is_in_A(x)+is_in_B(x)))
    }
    if(i==2){
      x<-rB()
      a<-(u <= 1/(is_in_A(x)+is_in_B(x)))
    }
    #if neither condition is selected, just run basic ar
    if(i==3){
      x<-runif(2)*4
      a<-(is_in_A(x)|is_in_B(x))
    }
  }
  return(x)
}

replicate(1000, bonf_ar_p9()) %>% t() %>% as.data.frame() %>% ggplot(aes(x=V1, y=V2))+geom_point()+xlim(c(0,4))+ylim(c(0,4))+xlab("x")+ylab("y")+theme_minimal()
```
Looks like my code is working!

b.

```{r}
set.seed(1234567)
points<-replicate(1000, bonf_ar_p9()) %>% t()
res<-cbind(points %>% apply(MARGIN=1, FUN=is_in_A), points %>% apply(MARGIN=1, FUN=is_in_B)) %>% rowSums()==2
tibble(
  mean=mean(res),
  se=sd(res)/sqrt(length(res))
)
```

My proportion of total draws that lies in $A \cap B$ (i.e., $\frac{m(A \cap B)}{m(A \cup B)}$) is $\boxed{0.246 \pm 0.014}$.
::::

::::: {.problem}
Consider the following probabilistic recursive algorithm for generating a geometric random variable with parameter \( p \).

```{r}
rgeo_onedraw <- function(p) {
  u <- runif(1)
  if (u < p) 
    return(1)
  else
    return(1 + rgeo_onedraw(p))
}
```

Show that this algorithm is locally correct for output \( G \) such that \( \prob(G = i) = (1 - p)^{i - 1} p \ind(i \in \{1, 2, 3, \ldots\}) \).
:::::

::::{.solution}
Suppose $G$ is output the`rgeo_onedraw` algorithm. 
Let $Y$ be the output of the recursive call.
We know \( \prob(Y = i) = (1 - p)^{i - 1} p \).
Prove original.

Let $G$ be the output of the call to the algorithm. For any $i$:
\[P(G=i)=P(G=X, X=i)+P(G=Y, Y=i)\]
\[P(G=i)=P(G=X | X=i) \cdot P(X=i) +P(G=Y | Y=i) \cdot P(Y=i)\]
\[P(G=i)= p \cdot P(X=i) + (1-p) \cdot P(Y=i)\]
\[P(G=i) - p \cdot P(X=i) = (1-p) \cdot P(Y=i)\]
Assuming that $P(G=i)$ and $P(X=i)$ are the same, because they are both based on the `rgeo_onedraw` algorithm:
\[(1-p) \cdot P(G=i)  = (1-p) \cdot P(Y=i)\]
\[P(G=i)  = P(Y=i) = (1 - p)^{i - 1} p\]
Because we supposed that $Y$ was an oracle. In short, we assumed the recursive call was true and used it to prove that the original algorithm was locally correct.
::::

