---
title: 'Homework #1:  First steps'
author: "Ethan Ashby"
output:
  html_document:
    css: ./homework-style.css
  pdf_document: 
    keep_tex:  TRUE
---

\newcommand{\mean}{\mathbb{E}}
\newcommand{\ind}{\mathbb{I}}
\newcommand{\real}{\mathbb{R}}
\newcommand{\unifdist}{\textsf{Unif}}
\newcommand{\expdist}{\textsf{Exp}}
\newcommand{\sd}{\operatorname{SD}}
\newcommand{\sqrtfcn}{\operatorname{sqrt}}

\renewcommand{\linethickness}{0.2mm}

```{r setup, include=FALSE, warning = FALSE, message= FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
#Color Format
colFmt = function(x,color){
  outputFormat = opts_knit$get("rmarkdown.pandoc.to")
  if(outputFormat == 'latex')
    paste("\\textcolor{",color,"}{",x,"}",sep="")
  else if(outputFormat == 'html')
    paste("<font color='",color,"'>",x,"</font>",sep="")
  else
    x
}
#Box Format
boxFmt = function(x){
  outputFormat = opts_knit$get("rmarkdown.pandoc.to")
  if(outputFormat == 'latex')
    paste("\\fbox{",x,"}")
  else if(outputFormat == 'html')
    paste("<span class=\"boxed\">",x,"</span>")
  else
    x
}
boxed <- boxFmt
```

Due:  Feb. 2, 2021

We will need the following libraries for this homework.
```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(readxl)
```

The following ensures that the same random choices are made each time the document is knit.
```{r}
set.seed(12345)
```


Instructions:  

1) Be sure to put the file `homework-style.css` in the parent directory of the directory where you have your .Rmd file.

1) Put your solutions between the lines that read `::::: {.solution}` and `:::::`.  

2) Knit the file to .html.

3) Print the file to .pdf using a browser.

4) Upload your .pdf file to Gradescope.

5) Mark the pages your problem solutions are on.

6) Be sure to box your work with the `\boxed{put in box}` $\LaTeX$ command.

7) Be sure to give nonintegral answers to four significant digits.

::::: {.problem}   
If $U \sim \unifdist([0, 1])$, what is $\mean[\sqrt{U}]$?
:::::

::::: {.solution}
Per **Fact 1** on the preclass notes for 1/25/21, if we have a random variable $X$ with density $f_x$, then $\mathbb{E}[g(x)]=\int_\mathbb{R} g(t) f_X(t)$. In this problem have a RV $U$ with density $\textrm{Unif}[0,1]$, implying that $\mathbb{E}[\sqrt{U}]={\displaystyle \int_{\mathbb{R}} \sqrt{x} \cdot \mathbb{I}(x \in [0,1]) dx}={\displaystyle \int_0^1 \sqrt{x} \: dx}=\bigg[\frac{2}{3}x^{3/2}\bigg]^1_0=\boxed{\frac{2}{3}}$.

To validate our answer empirically (using MC, we draw 1000 random values from $\unifdist([0, 1]$, calculate their square roots, and calculate the mean)
```{r}
set.seed(12345)
x<-runif(1000)
sqrtx<-sqrt(x)
paste("MC estimate:", round(mean(sqrtx), 4))
```
The Monte Carlo estimate is close to the analytical solution. I'm satisfied :)
:::::

::::: {.problem}
Suppose $\sd(X) = 3.2$ and $X_1, \ldots, X_{10}$ are iid as $X$.  What is the standard deviation of 
\[
\frac{X_1 + \cdots + X_{10}}{10}?
\]
:::::

::::: {.solution}
Per **Fact 4** on the preclass notes for 1/27/21, if $X_1, \cdots, X_{10}$ are iid $X$, then $SD \bigg( \frac{X_1 + \cdots + X_{10}}{10} \bigg)=\frac{SD(X)}{\sqrt{n}}$. This implies that $SD \bigg( \frac{X_1 + \cdots + X_{10}}{10} \bigg)=\frac{3.2}{\sqrt{10}}=\boxed{1.012}$
:::::

::::: {.problem}
Find a function $h$ such that $\mean[h(U)] = \int_0^1 \exp(\sqrt{x}) \ dx$ for $U$ a standard uniform random variable.
:::::

::::: {.solution}
Per **Fact 1** on Pre-class notes for 1/27/21, letting $h(u)=\frac{g(u)}{f_U(u)}$ (where $g(u)=\exp(\sqrt{u}) \: \mathbb{I}(u \in [0,1])$ is the function in integrand and $f_U(u)=\mathbb{I}(u \in [0,1])$ is the density of $U$) yields our desired result: $\mathbb{E}[h(u)]=\int g(u) \: du$. Thus, our desired $h(u)=\frac{\exp(\sqrt{u}) \: \mathbb{I}(u \in [0,1])}{ \mathbb{I}(u \in [0,1])}=\boxed{\exp(\sqrt{u}) \: \mathbb{I}(u \in [0,1])}$
:::::

::::: {.problem}
Continuing the last problem, write $R$ code to estimate $\int_0^1 \exp(\sqrt{x}) \ dx$ using 1000 draws.  Estimate the error and report your answer in the form $a \pm b$.
:::::

::::: {.solution}
```{r}
set.seed(12345)
#draw U's from standard uniform
U<-runif(1000)
#calculate expontiated sqrt U's
expsqrtU<-exp(sqrt(U))
#calculate arithmetic mean of these values, estimating expected value
paste("MC estimate:", round(mean(expsqrtU), 3))

#calculate sd of exponentiated sqrt U's
sigma<-sd(expsqrtU)
paste("Error:", round(sigma/sqrt(1000), 4))
```
The MC estimate for $\int_0^1 \exp(\sqrt{x}) \ dx = \boxed{2.022 \pm 0.014}$.
:::::

::::: {.problem}
For $T \sim \expdist(1)$ (so $T$ has density $f_T(t) = \exp(-t) \ind(t \geq 0))$, find a function $h$ such that 
\[
\mean[h(T)] = \int_0^\infty \exp(-x^{3 / 2}) \ dx.
\]
:::::

::::: {.solution}
Recall from problem 3 that our desired $h(T)=\frac{g(t)}{f_T(t)}$ where $g(t)$ is the integrand in the expected value expression and $f_T(t)$ is the density of $T$. Thus, $h(T)=\frac{\exp(-t^{3/2}) \: \mathbb{I}(t \in [0,\infty])}{\exp(-t) \: \mathbb{I}(t \in [0,\infty])}=\boxed{\exp(t-t^{3/2}) \: \mathbb{I}(t \in [0,\infty])}$
:::::

::::: {.problem}
Estimate $\int_0^\infty \exp(-x^{3 / 2}) \ dx$ using $T \sim \expdist(1)$ and importance sampling with $1000$ samples.
:::::

::::: {.solution}
```{r}
set.seed(12345)
#sample T's from exponential distribution with rate=1
ts<-rexp(1000, rate=1)
#transform the T's according to h(t) defined in the previous problem
trans_ts<-exp(ts-ts^(1.5))
paste("MC estimate:", round(mean(trans_ts), 4))
```

The Monte Carlo estimate for $\int_0^\infty \exp(-x^{3 / 2}) \ dx = \boxed{0.9237}$

:::::

::::: {.problem}
Consider the integral
\[
a = \int_{x = 0}^\infty \exp(-x^{3 / 2}) \ dx.
\]
Use $t = 1 - 1 / (1  + x)$ to show that
\[
a = \int_{t = 0}^1 \exp(-(t / (1 - t))^{3/2})/ (1 - t)^2 \ dt.
\]
::::: 

::::: {.solution}
First, let's write $t=1-1/(1+x)$ in terms of $x$. $t=1-1/(1+x) \rightarrow  1 = \frac{1}{t} - \frac{1}{t \: (1+x)} \rightarrow (1+x)=\frac{1+x}{t}-\frac{1}{t} \rightarrow \frac{t-1}{t} (1+x)=-\frac{1}{t} \rightarrow x =- \frac{1}{t-1} -1 = \frac{-t}{t-1}=\frac{t}{1-t}$.
Thus, we've shown that $x=\frac{t}{1-t}$.
Note that then $dx=\frac{1}{(t-1)^2} \ dt$ using the quotient rule.
Now we can substitute in $\frac{t}{1-t}$ for $x$ and $\frac{1}{(t-1)^2} \ dt$ for $dx$ in the integral. But we must also change the limits of integration. In the original integral, $x \in [0, \infty)$ and since $t=1-1/(1+x)$, $t \in [0, 1)$.
Thus, our final integral, where we've changed the integrand, limits, and differential is as expected:
$a = \int_{t = 0}^1 \exp(-(t / (1 - t))^{3/2})/ (1 - t)^2 \ dt$.
:::::

::::: {.problem}
Continuing the last problem, use 1000 standard uniforms to estimate $a$ using the transformed integral.  Write your answer in the form $a \pm b$.
:::::

::::: {.solution}
```{r}
set.seed(12345)
#random uniform ts
tunifs<-runif(1000)
#transformed ts
t_trans<-exp(-(tunifs/(1-tunifs))^1.5)/(1-tunifs)^2
paste("MC estimate:", mean(t_trans))
paste("Est. error:", sd(t_trans)/sqrt(1000))
```

MC estimate for the value of the transformed integral is $\boxed{0.8751 \pm 0.0200}$
:::::

::::: {.problem}
Using 1000 samples from uniforms over $[0, 1]^3$, estimate the value of 
\[
\int_{(x, y, z) \in [0, 1]^3} \sqrt{x + y}\exp(-x - z) \ d\real^3.
\]
:::::

::::: {.solution}
This is where MC solutions get really good!
```{r}
set.seed(12345)
#generate 1000 by 3 matrix containing random x,y,z trios over [0,1]
xyz<-replicate(3, runif(1000))
#create function that calculates h(x,y,z)
fun<-function(vec){sqrt(vec[1]+vec[2])*exp(-vec[1]-vec[3])}
#apply this function over every row of xyz, i.e. 1000 instances of x,y,z trios
Ts<-apply(xyz, MARGIN=1, FUN = fun)
paste("MC estimate:", round(mean(Ts),4))
```
My Monte Carlo estimate for the integral is $\boxed{0.3751}$.
:::::

::::: {.problem}
Using 1000 samples, estimate the following integral:
\[
\int_{(x_1, \ldots, x_{20}) \in [0,1]^{20}} \min\left(11, \sum_{i = 1}^n x_i\right) \ d\real^{20}.
\]
[Note:  these types of *minimum* integrands arise often in mathematical finance.]
:::::

::::: {.solution}

```{r}
set.seed(12345)
#generate 1000 by 20 matrix containing random (x1, ..., x20) from [0,1]^20
inputs<-replicate(20, runif(1000))
#create function that calculates the integrand function
fun<-function(vec){min(11, sum(vec))}
#apply this function over every row of inputs, i.e. 1000 instances of inputs
Ts<-apply(inputs, MARGIN=1, FUN = fun)
paste("MC estimate:", round(mean(Ts),4))
```
My MC estimate for the value of the integral $\boxed{9.8157}$
:::::
